Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers

### 2023-04-20 2_SSL_EST/article/article_draft_est.tex

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/history_efficacy.pdf}
 \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history. Data points indicate best performers by year of market introduction. Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Information, Section 1 for details). Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Information, Section 4 for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure}

\begin{figure}[h!]
\centering
  \includegraphics[height=6.5cm]{2_SSL_EST/article/figures/cost_lamp_small.pdf}
  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{council2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs. Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications. Data on CFL sales prices collected from Eger \& Ehlhardt \cite{eger2018origin} as well as various consumer watchdog databases and publications. Incandescent light bulb price is assumed constant based on the average in the covered time period. See Supplementary Information, Section 4, for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

@@

\clearpage
\begin{figure}[h!]
 \centering
 \includegraphics[height=15cm]{2_SSL_EST/article/figures/chip_architecture_overview.pdf}
 \caption{Historical evolution of light-emitting diode chip architectures. Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture. Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture. The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip. Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity. Years indicated correspond to the earliest identified patent priority date. Black frames around certain designs indicate chip designs not brought to large-scale production. Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well. Source: adapted and compiled from multiple patents and industry publications. See Supplementary Information, Section 4, for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
\end{figure}

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section 1.3 of the Supplementary Information. Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy. The corresponding LED architectures and associated down-conversion phosphors are shown for reference: 2003 - flip-chip with YGAG phosphor; 2010 - flip-chip with 258 phosphor; 2016 - flip-chip with PFS phosphor; 2020 - flip-chip with SALON phosphor. A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}. Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and invention history provided in Section 3 in the Supplementary Information. Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010. Abbreviations: Scat. = Scattering. Source (Efficiency): own elaboration based on data in Figure SI11. Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_efficiency.pdf}
 \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA. Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for the four years used in Figure \ref{fgr:waterfall}: 2003, 2010, 2016, and 2020. To provide additional historical context, data for 1997 is included for selected sub-efficiencies. "N/A" denotes sub-efficiencies where 1997 data could not be calculated to different reasons: $\eta_{Vf}$, $\eta_{droop}$ depend on the device current, which was below 350mA in 1997, making comparison with contemporary devices difficult. $\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures light extraction efficiency no longer depends on ITO, see Figure \ref{fgr:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column. $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.  Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}. Source: own elaboration based on data represented in Figure \ref{fgr:waterfall} and Figure SI11 in the Supplementary Information, as detailed in Section Methods and Data Collection.}
 \label{fgr:breakthroughs_efficiency}
\end{figure}

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_consumer-experience.pdf}
 \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs. Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year. Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 11 in Section 3 of the Supplementary Information. Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison. Horizontal lines represent typical colour temperatures of “traditional” light sources, shown for reference. The desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures. Horizontal red arrow indicates desirable higher values of colour rendering index (CRI). LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG - Nichia, 1998 \cite{bando1998development}; YGAG - Lumileds, 2002 \cite{Mueller2002}; 258 - Lumileds, 2005 \cite{MuellerMach2005}; SLA - Lumileds, 2014 \cite{Pust2014}; PFS - GE, 2015 \cite{Murphy2015}; QD - Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON - Osram, 2019 \cite{Hoerder2019}.}
 \label{fgr:consumer_experience}
\end{figure}

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17.5cm]{2_SSL_EST/article/figures/costmodel_results_years.pdf}
 \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S. location. Consecutive steps shown on the x-axis. Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020. Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process. Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020. For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information. Abbreviations: Litho – Lithographic Process, Insp. – Inspection, Depn. – Deposition, CMP - Chemical-Mechanical Planarization}
 \label{fgr:costmodel}
\end{figure*}

>>>>

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
 \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history. Data points indicate best performers by year of market introduction. Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Information, Section 1 for details). Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data. The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al. \cite{Murphy2012}. The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year. For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines. Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot. Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal. - Halogen, Incd. - Incandescent. Source: own synthesis of published data based on a visual approach proposed by Azevedo et al. \cite{azevedo2009transition}. See Supplementary Information, Section 4 for the full list of sources and references.}
 \label{fgr:history_efficacy}
\end{figure}

\begin{figure}[h!]
\centering
  \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020. Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S. Department of Energy (DOE) \cite{council2013assessment}. Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs. Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications. Data on CFL sales prices collected from Eger \& Ehlhardt \cite{eger2018origin} as well as various consumer watchdog databases and publications. Incandescent light bulb price is assumed constant based on the average in the covered time period. See Supplementary Information, Section 4, for the full list of sources and references.}
  \label{fgr:cost_lamp_small}
\end{figure}

@@

\clearpage
\begin{figure}[h!]
 \centering
 \includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
 \caption{Historical evolution of light-emitting diode chip architectures. Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture. Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture. The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip. Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity. Years indicated correspond to the earliest identified patent priority date. Black frames around certain designs indicate chip designs not brought to large-scale production. Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well. Source: adapted and compiled from multiple patents and industry publications. See Supplementary Information, Section 4, for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
\end{figure}

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section 1.3 of the Supplementary Information. Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy. The corresponding LED architectures and associated down-conversion phosphors are shown for reference: 2003 - flip-chip with YGAG phosphor; 2010 - flip-chip with 258 phosphor; 2016 - flip-chip with PFS phosphor; 2020 - flip-chip with SALON phosphor. A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}. Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and invention history provided in Section 3 in the Supplementary Information. Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010. Abbreviations: Scat. = Scattering. Source (Efficiency): own elaboration based on data in Figure SI11. Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
 \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA. Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for the four years used in Figure \ref{fgr:waterfall}: 2003, 2010, 2016, and 2020. To provide additional historical context, data for 1997 is included for selected sub-efficiencies. "N/A" denotes sub-efficiencies where 1997 data could not be calculated to different reasons: $\eta_{Vf}$, $\eta_{droop}$ depend on the device current, which was below 350mA in 1997, making comparison with contemporary devices difficult. $\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures light extraction efficiency no longer depends on ITO, see Figure \ref{fgr:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column. $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.  Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}. Source: own elaboration based on data represented in Figure \ref{fgr:waterfall} and Figure SI11 in the Supplementary Information, as detailed in Section Methods and Data Collection.}
 \label{fgr:breakthroughs_efficiency}
\end{figure}

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
 \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs. Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year. Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 11 in Section 3 of the Supplementary Information. Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison. Horizontal lines represent typical colour temperatures of “traditional” light sources, shown for reference. The desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures. Horizontal red arrow indicates desirable higher values of colour rendering index (CRI). LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG - Nichia, 1998 \cite{bando1998development}; YGAG - Lumileds, 2002 \cite{Mueller2002}; 258 - Lumileds, 2005 \cite{MuellerMach2005}; SLA - Lumileds, 2014 \cite{Pust2014}; PFS - GE, 2015 \cite{Murphy2015}; QD - Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON - Osram, 2019 \cite{Hoerder2019}.}
 \label{fgr:consumer_experience}
\end{figure}

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
 \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S. location. Consecutive steps shown on the x-axis. Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020. Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process. Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020. For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information. Abbreviations: Litho – Lithographic Process, Insp. – Inspection, Depn. – Deposition, CMP - Chemical-Mechanical Planarization}
 \label{fgr:costmodel}
\end{figure*}
